A Bayesian ranking-statistic for binary black holes mergers in the second LIGO observing run

### 2021-05-28 main.tex
complete draft

% Keywords
\newcommand{\bilby}{{\sc \href{https://lscsoft.docs.ligo.org/bilby/}{\texttt{Bilby}}}\xspace}
\newcommand{\bilbypipe}{{\sc \texttt{bilby\_pipe}}\xspace}
\newcommand{\dynesty}{{\sc \texttt{dynesty}}\xspace}
\newcommand{\cpnest}{{\sc cpnest}\xspace}
\newcommand{\ptemcee}{{\sc ptemcee}\xspace}
\newcommand{\gwpy}{{\sc \href{https://gwpy.github.io/}{\texttt{GWpy}}}\xspace}
\newcommand{\imrphenomp}{{\sc IMRPhenomPv2}\xspace}
\newcommand{\seob}{{\sc SEOBNRv4PHM}\xspace}

@@

\begin{abstract}
The detection of an intermediate-mass black hole population ($10^2-10^6$~\msun) will provide probes to their formation environments (e.g., disks of active galactic nuclei, globular clusters) and illuminate a potential pathway to produce supermassive black holes. Numerous methods to search for $10^4-10^6\ \msun$ intermediate-mass black holes exist, while the primary method to find $10^2-10^3\ \msun$ is by identifying gravitational-waves emitted from the merger of stellar binaries. Ground-based gravitational-wave detectors are in principle sensitive to such mergers and have found one $142^{+28}_{-16}\ \msun$ intermediate-mass black hole. In reality, ground-based detector data contain numerous short-duration instrumental transients that mimic the gravitational-wave signals from mergers capable of producing intermediate-mass black holes, making intermediate-mass merger detection challenging. Here we demonstrate a ranking statistic utilizing Bayesian inference to detect binary black hole mergers with a total mass $>55$~\msun. \avi{Say why it improves on prev techniques}. We apply this technique to candidate events with total masses $>55$~\msun during LIGO's second observing run to search for previously unresolved gravitational-wave signals from this heavy category of binary black holes. Our analysis does not discover new intermediate-mass black holes. However, we find support for some stellar-mass binary black holes unreported in GWTC-1.
\end{abstract}

@@

Although a significant fraction of the glitches can be removed by testing them for coherence amongst various ground based detectors and performing matched-filtering, these methods are insufficient to remove all the glitches~\cite{ pycbc_short_duration_transients, pe_with_glitch, blip_glitches}. One method to account for more glitches while searching for CBC gravitational-waves is to rely on a Bayesian odds~\cite{bci, kanner2016leveraging, BCR1, BCR2, bcr_gw151216, bayesian_odds}. In this paper, we rank O2's candidate gravitational-wave signals from $55-500\ \msun$ total mass systems using a Bayesian method called the Bayesian Coherence Odds-Ratio \bcr~\cite{BCR1}. \avi{Should explain: whats the advantage of this? Why is it useful for triggers with $M_T>55\msun$?} The \bcr used in this study is a bootstrap-Bayesian odds calculated using the probability of data under the hypothesis that the data contains coherent signals vs. incoherent glitches and empirically calculated prior-odds that describe how likely each hypothesis is. After computing the candidate \bcr values, we calculate the candidate's probability of astrophysical origins, \pastrobcr. Finally, if the candidate has a \pastro reported by the the LIGO-Virgo-KAGRA (LVK) collaboration in \GWTC~\cite{GWTC1}, the \pycbc-team~\cite{pycbc_code, pycbc_og0, pycbc_og1, pycbc_og2, pycbc_og3, pycbc_og4, pycbc_og5, pycbc_og6, pycbc_single_det, pycbc_ogc_2}, by the Institute of Advanced study's team (\IAS)~\cite{IAS0, IAS1, IAS2}, or by \citet{bayesian_odds}, we compare their \pastro with the \pastrobcr.  We highlight several innovations unique to this work. First, we use our Bayesian framework to analyze all candidate signals and background triggers (incoherent signal-free data) with total masses above $> 55 \msun$ in O2. Second, we empirically tune prior-odds for finding a signal or a glitch in O2 using \bcr distributions for background triggers and simulated signals. Finally,  we use the tuned-\bcr background distribution to compute how significant candidate signals are from the background.

We find that (a) high-mass events reported in the \GWTC, including GW170729 (the heaviest event in \GWTC) are very statistically significant $\pastrobcr>0.9$; (b) three out of the eight \IAS events and candidates have $\pastrobcr>0.5$, corroborating \IAS's detection claims for those events (GW170304, GW170727, GW170817A); and that (c) our ranking statistic does not identify any new intermediate-mass black holes, but does identify an unreported marginal stellar-mass binary black hole candidate, 170222 with $\pastrobcr\sim0.5$ \avi{Need to doublecheck that this wasnt in the PyCBC sub-threshold paper}.

@@

The standard framework to identify CBC gravitational-wave signals in data is to quantify the significance of candidates with null-hypothesis significance testing~\cite{GWTC1, GWTC2}. In this framework, the candidates' ranking statistic is compared against a background distribution in a frequentist approach. The independent matched-filter searches, e.g. \pycbc~\cite{pycbc_og4}, \spiir~\cite{spiir} and \gstlal~\cite{sachdev2019gstlal}, and the coherent burst search \cwb~\cite{cwb} used by LVK to search for signals in gravitational-wave data all use ranking statistics in such a manner~\cite{GWTC1}. \avi{Holdup: isnt GSTLAL's method a bootstrapped-Bayesian method?} Both \pycbc and \gstlal's ranking statistic incorporate information of the relative likelihood that the data contains a coherent signal versus just noise. In contrast, \cwb's ranking statistic uses the information of coherent energy present in the network of detectors~\cite{GWTC1}.

\avi{I discuss the benefits of the BCR here...}
Bayesian inference offers an alternative means to rank the significance of candidate events by computing the odds that the data contain a transient gravitational-wave signal vs. instrumental glitches or noise~\cite{BCR1}. This method relies on accurate models for the signal and glitch morphologies and populations~\cite{BCR1}. In principle, Bayesian odds is the optimal method for hypothesis testing~\cite{BCR2}. Much of its power comes from the Bayesian ``evidence'': a marginalized likelihood. The evidence is the correct likelihood of having obtained the data given a hypothesis. However, due to the high computational cost to calculate the evidence, it is not explicitly used in current matched filter searches. Here, we explore a middle-ground hybrid between frequentist matched filtered searches and Bayesian odds. We compute Bayesian evidences for foreground and background data under the assumption that they either contain coherent gravitational-wave signals, noise, or glitches. However, instead of computing true Bayesian odds, we use the evidences to form a bootstrapped distribution for the odds of foreground and background events to form a frequentist ranking statistic.

@@

\subsection{Estimation of astrophysical signal probability}
Candidate \bcr-values are either statistically insignificant compared to the background \bcr distribution, implying the candidate is more probable to be an incoherent instrumental feature (the $\mathcal{H}_I$ null-hypothesis), or statistically significant relative to the background distribution, indicating the possible presence of an astrophysical signal (the $\mathcal{H}_S$ hypothesis). A false alarm probability with trial factors \avi{(need to define a trial factor)}, \fap, for the candidate \bcr-value can quantify the significance. The \fap is the probability that a candidate originating from a non-astrophysical source can be falsely identified as a signal.

To calculate the \fap, each candidate \bcr is considered a single statistical trial that can occur at a fixed false alarm probability $f$, where $f$ is the probability of observing a background $\bcr'$ greater than or equal to the candidate \bcr,
\begin{equation}
    f = \frac{\text{Count of } \bcr' \leq \bcr}{\text{Count of } \bcr'} \ .
\end{equation}
The false alarm probability with trials \fap that the \bcr measurement occurs at least once for $N$ trials $(N > 0)$, where $N$ is the number of candidate signals, is \avi{(This is very convoluted)}
\begin{equation}
    \fap = 1 - (1-f)^N \ .
\end{equation}

@@

\avi{Should I mention that we need \pycbc  as a pre-selection step to computing? }

@@

In addition to finding candidate events, \pycbc also identified several million background triggers for each time-frame, by searching background data manufactured by time-sliding data within that time-frame. The background triggers help quantify the candidate events' significance for the respective time-frames. Finally, to test the search's sensitivity, \pycbc produced and searched for thousands of simulated signals. \avi{Do i need to contextualize this choice more, ie talk about what would happen if I chose a narrower range?}

@@

}
% time for chunk 14
% 1176955218, Apr 23, 2017, 4:00 UTC
% 1178294418, May 08, 2017, 16:00 UTC
\caption[ BCR search space.]{The template bank (pink) used by \pycbc to search a section of O2 data from $\text{April 23 - May 8, 2017}$. Our search is constrained to the parameter space enclosed by the dashed line. The candidate, background and simulated triggers detected in this region of the parameter space during this period are plotted in orange, gray and blue respectively.}\label{fig:templateBank}
\end{figure*}

@@

The waveform template we utilize is \imrphenomp, a phenomenological waveform template constructed in the frequency domain that models the in-spiral, merger, and ring-down (IMR) of a compact binary coalescence~\citep{khan2016frequency}. Although there exist gravitational-wave templates such as IMRPhenomPv3HM, NRSur7dq4, \seob~\cite{seobnrv4phm} (and others) which incorporate more physics, such as information on higher-order modes, we use \imrphenomp as it is computationally inexpensive compared to others.

>>>>

% Keywords
\newcommand{\bilby}{{\sc \href{https://lscsoft.docs.ligo.org/bilby/}{\texttt{Bilby}}}\xspace}
\newcommand{\bilbypipe}{{\sc \texttt{bilby\_pipe}}\xspace}
\newcommand{\dynesty}{{\sc \texttt{dynesty}}\xspace}
\newcommand{\cpnest}{{\sc cpnest}\xspace}
\newcommand{\ptemcee}{{\sc ptemcee}\xspace}
\newcommand{\gwpy}{{\sc \href{https://gwpy.github.io/}{\texttt{GWpy}}}\xspace}

\newcommand{\imrphenomp}{{\sc \texttt{IMRPhenomPv2}}\xspace}
\newcommand{\seob}{{\sc \texttt{SEOBNRv4PHM}}\xspace}
\newcommand{\nrsur}{{\sc \texttt{NRSur7dq4}}\xspace}
\newcommand{\imrxhm}{{\sc \texttt{IMRPhenomXPHM}}\xspace}

@@

\begin{abstract}
The detection of an intermediate-mass black hole population ($10^2-10^6$~\msun) will provide probes to their formation environments (e.g., disks of active galactic nuclei, globular clusters) and illuminate a potential pathway to produce supermassive black holes. Numerous methods to search for $10^4-10^6\ \msun$ intermediate-mass black holes exist, while the primary method to find $10^2-10^3\ \msun$ is by identifying gravitational-waves emitted from the merger of stellar binaries. Ground-based gravitational-wave detectors are in principle sensitive to such mergers and have found one $142^{+28}_{-16}\ \msun$ intermediate-mass black hole. In reality, ground-based detector data contain numerous short-duration instrumental transients that mimic the gravitational-wave signals from mergers capable of producing intermediate-mass black holes, making intermediate-mass merger detection challenging. Here we demonstrate a ranking statistic utilizing Bayesian inference to detect binary black hole mergers with a total mass $>55$~\msun. We apply this technique to candidate events with total masses $>55$~\msun during LIGO's second observing run to search for previously unresolved gravitational-wave signals from this heavy category of binary black holes. Our analysis does not discover new intermediate-mass black holes. However, we find support for some stellar-mass binary black holes unreported in GWTC-1.
\end{abstract}

% \avi{Say why it improves on prev techniques}

@@

Although a significant fraction of the glitches can be discriminated by testing them for coherence amongst various ground based detectors and performing matched-filtering, these methods are insufficient to identify all glitches~\cite{ pycbc_short_duration_transients, pe_with_glitch, blip_glitches}. One method to discriminate more glitches while searching for CBC gravitational-waves is to rely on a Bayesian odds~\cite{bci, kanner2016leveraging, BCR1, BCR2, bcr_gw151216, bayesian_odds}. Utilising Bayesian odds is the optimal method for discriminating glitches from signals~\cite{BCR2}. In this paper, we rank O2's candidate gravitational-wave signals from $55-500\ \msun$ total mass systems using a Bayesian method called the Bayesian Coherence Ratio \bcr~\cite{BCR1}. The \bcr used in this study is a bootstrap-Bayesian odds that estimates the true Bayesian odds. The \bcr is calculated using the probability of data under the hypothesis that the data contains coherent signals vs. incoherent glitches and empirically calculated prior-odds that describe how likely each hypothesis is. After computing the candidate \bcr values, we calculate the candidate's probability of astrophysical origins, \pastrobcr. Finally, if the candidate has a \pastro reported by the the LIGO-Virgo-KAGRA (LVK) collaboration in \GWTC~\cite{GWTC1}, the \pycbc-team~\cite{pycbc_code, pycbc_og0, pycbc_og1, pycbc_og2, pycbc_og3, pycbc_og4, pycbc_og5, pycbc_og6, pycbc_single_det, pycbc_ogc_2}, by the Institute of Advanced study's team (\IAS)~\cite{IAS0, IAS1, IAS2}, or by \citet{bayesian_odds}, we compare their \pastro with the \pastrobcr.  We highlight several innovations unique to this work. First, we use our Bayesian framework to analyze all candidate signals and background triggers (incoherent signal-free data) with total masses above $> 55 \msun$ in O2. Second, we empirically tune prior-odds for finding a signal or a glitch in O2 using \bcr distributions for background triggers and simulated signals. Finally,  we use the tuned-\bcr background distribution to compute how significant candidate signals are from the background.

We find that (a) high-mass events reported in the \GWTC, including GW170729 (the heaviest event in \GWTC) are very statistically significant $\pastrobcr>0.9$; (b) three out of the eight \IAS events and candidates have $\pastrobcr>0.5$, corroborating \IAS's detection claims for those events (GW170304, GW170727, GW170817A); and that (c) our ranking statistic does not identify any new intermediate-mass black holes, but does identify an unreported marginal stellar-mass binary black hole candidate, 170222 with $\pastrobcr\sim0.5$ %\avi{Need to doublecheck that this wasnt in the PyCBC sub-threshold paper}.

@@

The standard framework to identify CBC gravitational-wave signals in data is to quantify the significance of candidates with null-hypothesis significance testing~\cite{GWTC1, GWTC2}. In this framework, the candidates' ranking statistic is compared against a background distribution. The independent matched-filter searches, e.g. \pycbc~\cite{pycbc_og4}, \spiir~\cite{spiir} and \gstlal~\cite{sachdev2019gstlal}, and the coherent burst search \cwb~\cite{cwb} used by LVK to search for signals in gravitational-wave data all use ranking statistics in such a manner~\cite{GWTC1}. Both \pycbc and \gstlal's ranking statistic incorporate information of the relative likelihood that the data contains a coherent signal versus just noise. In contrast, \cwb's ranking statistic uses the information of coherent energy present in the network of detectors~\cite{GWTC1}.

Bayesian inference offers an alternative means to rank the significance of candidate events by computing the odds that the data contain a transient gravitational-wave signal vs. instrumental glitches or noise~\cite{BCR1}. This method relies on accurate models for the signal and glitch morphologies and populations~\cite{BCR1}. In principle, Bayesian odds is the optimal method for hypothesis testing~\cite{BCR2}. Much of its power comes from the Bayesian ``evidence'': a marginalized likelihood. The evidence is the correct likelihood of having obtained the data given a hypothesis. However, due to the high computational cost to calculate the evidence, it is not explicitly used in current matched filter searches. Here, we explore a middle-ground hybrid between frequentist matched filtered searches and Bayesian odds. We compute Bayesian evidences for foreground and background data under the assumption that they either contain coherent gravitational-wave signals, noise, or glitches. However, instead of computing true Bayesian odds, we use the evidences to form a bootstrapped distribution for the odds of foreground and background events to form a frequentist ranking statistic.

@@

\subsection{Estimation of astrophysical signal probability}
Candidate \bcr-values are either statistically insignificant compared to the background \bcr distribution, implying the candidate is more probable to be an incoherent instrumental feature (the $\mathcal{H}_I$ null-hypothesis), or statistically significant relative to the background distribution, indicating the possible presence of an astrophysical signal (the $\mathcal{H}_S$ hypothesis). A false alarm probability with trial factors, \fap, for the candidate \bcr-value can quantify the significance. The \fap for a \bcr-value is the probability that a candidate originating from a non-astrophysical source can be falsely identified as a signal once in $N$ trials, and is given by 
\begin{equation}
    \fap = 1 - (1-f)^N \ ,
\end{equation}
where $f$ is the probability of observing a background $\bcr'$ greater than or equal to the candidate \bcr,
\begin{equation}
    f = \frac{\text{Count of } \bcr' \leq \bcr}{\text{Count of } \bcr'} \ .
\end{equation}

@@

% \avi{Should I mention that we need \pycbc  as a pre-selection step to computing? }

@@

In addition to finding candidate events, \pycbc also identified several million background triggers for each time-frame, by searching background data manufactured by time-sliding data within that time-frame. The background triggers help quantify the candidate events' significance for the respective time-frames. Finally, to test the search's sensitivity, \pycbc produced and searched for thousands of simulated signals. 
% \avi{Do i need to contextualize this choice more, ie talk about what would happen if I chose a narrower range?}

@@

}
% time for chunk 14
% 1176955218, Apr 23, 2017, 4:00 UTC
% 1178294418, May 08, 2017, 16:00 UTC
\caption[ BCR search space.]{The templates (pink) used by \pycbc to search a section of O2 data from $\text{April 23 - May 8, 2017}$. Our search is constrained to the parameter space enclosed by the dashed line. The candidate, background and simulated triggers detected in this region of the parameter space during this period are plotted in orange, gray and blue respectively.}\label{fig:templateBank}
\end{figure*}

@@

The waveform template we utilize is \imrphenomp, a phenomenological waveform template constructed in the frequency domain that models the in-spiral, merger, and ring-down (IMR) of a compact binary coalescence~\citep{khan2016frequency}. Although there exist gravitational-wave templates such as \imrxhm~\cite{imrphenompxhm}, \nrsur~\cite{nrsur7dq4} and \seob~\cite{seobnrv4phm} which incorporate more physics, such as information on higher-order modes, we use \imrphenomp as it is computationally inexpensive compared to others.
